miniKanren with fair search strategies

### 2019-05-02 Paper/paper.tex
tuning

\section{introduction}
\NOTE[MVC]{We need clear definitions of the following: fair, almost-fair, and unfair. Either use bullet points, or definition environment. Google ``latex math definition.''}
\NOTE[MVC]{A quick intro to miniKanren would be nice here. We should mention The Reasoned Schemer, 2nd and say DFSi is used there.}

@@

Given the following program, it is natural to expect lists of each letter
constitute $1/4$ in the answer. However, DFS$_\textrm{i}$, the current search
strategy, results in much more lists of \texttt{a}s than lists of other
letters. And some letters, e.g. \texttt{c} and \texttt{d}, are rarely seen.
The situation would be exacerbated if \conde contains more clauses.

@@

Now we are in a middle place between fair and unfair -- search strategies
with  \emph{almost-fair} \disj{} should allocate resources so evenly among
disjunctive  goals\citep{seres1999algebra} that the the maximal ratio of
resources is bounded by a constant. Our new search strategy,
DFS$_\textrm{bi}$, has almost-fair \disj{}. It is
fair when the number of goals is a power of 2, otherwise, some goals are 
allocated twice more resources than the others. In the previous example,
DFS$_\textrm{bi}$
gives the same result. And in the following example, where the \conde has 5 
clause, the clauses of \texttt{b}, \texttt{c}, and \texttt{d} are allocated 
more resources.

@@

A more interesting situation is when the first conjunctive goal produces 
infinite many answers. Consider the following example, a naive specification of 
fair \conj{} might require search strategies to produce all sorts of singleton 
lists, but no longer ones, which makes the strategies \emph{incomplete}.

@@

Our solution is requiring a search strategy with \emph{fair} \conj{} to 
package answers in bags, where each bag contains finite answers, and to allocate 
resources evenly among search spaces derived from answers in the same bag. The 
way to package depends on search strategy. And how to allocate resources among 
search space related to different bags is unspecified. Our definition of fair 
\conj{} is orthogonal with completeness. For example, a naively fair strategy 
is fair but not complete, while BFS is fair and complete.

BFS packages answers by their costs. The \emph{cost} of a answer is its 
depth in the search tree (i.e. the number of calls to relational definitions 
required to find them) \citep{seres1999algebra}. In the following example, the 
cost of an answer is equal to the length of the inner lists plus the length of 
the outer list. Noted that BFS produces answers in increasing order of cost.

@@

Balanced interleaving DFS (DFS$_\textrm{bi}$) has almost-fair \disj{} and unfair \conj{}. 
The implementation of DFS$_\textrm{bi}$ differs from DFS$_\textrm{i}$ in the \disj{} macro. We list the 
new \disj{} with its helpers in Fig.~\ref{balanced-disj}. The first helper 
function, \texttt{split}, takes a list of goals \texttt{ls} and a procedure 
\texttt{k}, partitions \texttt{ls} into two sub-lists of roughly equal length, 
and returns the application of \texttt{k} to the two sub-lists. \texttt{disj*} 
takes a non-empty list of goals \texttt{gs} and returns a goal. With the help 
of \texttt{split}, it essentially constructs a \emph{balanced} binary tree 
where leaves are elements of \texttt{gs} and nodes are \texttt{disj2}, whence 
the name of this search strategy. In contrast, the \disj{} in DFS$_\textrm{i}$ essentially 
constructs the same sort of binary tree in one of the most unbalanced forms.

@@

As we mentioned in section 2.2, our BFS bag answers by their cost. The bagging 
information is recorded subtly -- the \texttt{car}s of a search space have cost 
0 (i.e. they are in the same bag), and the costs of answers in thunk are 
computed recursively then increased by one. It is even more subtle that
\texttt{append-inf/fair} and the \texttt{append-map-inf/fair} respects the cost 
information. We make these facts more obvious by changing the type of search 
space, modifying related function definitions, and introducing a few more 
functions.

@@

\subsection{comparison with Silvija's BFS}

In this section, we compare the pure subset of our optimized BFS with the BFS 
found in \citep{seres1999algebra}. We focus on the pure subset because 
Silvija's system is pure. Her system represents search spaces with streams of 
lists of answers, where each list is a bag.

To compare efficiency, we translate her Haskell code into Racket (See 
supplements for the translated code). The translation is fairly straightforward 
due to the similarity in both logic programming system and search space 
representation. The translated code is longer and slower. Details about 
efficiency difference are in section 6.

We prove the two search strategies are equivalent in Coq. Since search space 
can be infinite, we should use a co-inductive data type. However, Coq is too 
strict in the guardedness condition to accept a direct translation of the 
implementations. Therefore, we prove core theorems with finite search space 
instead. In order to generalize the conclusion to the cases with infinite 
search space, we prove a few more theorems saying that whenever we query 
answers lower than some finite cost, we can restrict goals to truncate search 
spaces at some finite depth without changing the query result. (See supplements 
for the formal proof)

@@

In this section, we compare the efficiency of search strategies. A concise 
description is in Table~\ref{compare-efficiency}. A hyphen means running out of 
memory. The first two benchmarks are taken from 
\citep{friedman_reasoned_2018}. \texttt{reverso} is from 
\citep{rozplokhas2018improving}. Next two benchmarks 
about quine are modified from a similar test case in \citep{byrd2017unified}. 
The modifications are made 
to circumvent the need for symbolic constraints (e.g. $\neq$, 
\texttt{absento}). Our version generates de 
Bruijnized expressions and prevent closures getting into list. The two 
benchmarks differ in the \conde clause order of their relation interpreters. 
The last two 
benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love 
you)}. This benchmark is also inspired by \citep{byrd2017unified}. Again, the 
sibling benchmarks differ in the \conde clause order of their relation 
interpreters. The first one 
has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the 
end, while the other has them at the beginning. We conjecture that DFS$_\textrm{i}$ would 
perform badly in the second case because elimination rules complicate the 
problem when running backward. The evaluation supports our conjecture.

>>>>

\section{introduction}
\NOTE[MVC]{We need clear definitions of the following: fair, almost-fair, and unfair. Either use bullet points, or definition environment. Google ``latex math definition.''}
\NOTE[MVC]{A quick intro to miniKanren would be nice here. We should mention The Reasoned Schemer, 2nd and say DFSi is used there.}
\NOTE[MVC]{We should cite the monad paper in introduction.}

@@

Given the following program, it is natural to expect lists of each letter
constitute $1/4$ in the answer. DFS$_\textrm{i}$, the current search
strategy, however, results in many more lists of \texttt{a}s than lists
of other letters. And some letters, e.g. \texttt{c} and \texttt{d}, are
rarely seen. The situation would be exacerbated if \conde contains more clauses.

@@

Now we are in a middle place between fair and unfair -- search strategies
with  \emph{almost-fair} \disj{} should allocate resources so evenly among
disjunctive  goals\citep{seres1999algebra} that the the maximal ratio of
resources is bounded by a constant. Our new search strategy,
DFS$_\textrm{bi}$, has almost-fair \disj{}. It is
fair when the number of goals is a power of 2, otherwise, some goals are 
allocated twice as many resources than the others. In the previous example,
DFS$_\textrm{bi}$
gives the same result. And in the following example, where the \conde has 5 
clause, the clauses of \texttt{b}, \texttt{c}, and \texttt{d} are allocated 
more resources.

@@

A more interesting situation is when the first conjunctive goal produces 
infinite many answers. Consider the following example, a naive specification of 
fair \conj{} might require search strategies to produce all sorts of singleton 
lists, but no longer ones, which makes the strategies \emph{incomplete}. 
\NOTE[MVC]{incomplete w.r.t. what? where's the definition of incomplete?}

@@

Our solution is requiring a search strategy with \emph{fair} \conj{} to 
package answers in bags, where each bag contains finite answers, and to allocate 
resources evenly among search spaces derived from answers in the same bag. The 
way to package depends on search strategy. And how to allocate resources among 
search space related to different bags is unspecified. Our definition of fair 
\conj{} is orthogonal with completeness. For example, a naively fair strategy 
is fair but not complete, while BFS is fair and complete. 
\NOTE[MVC]{Also here, complete w.r.t what?}

BFS packages answers by their costs. The \emph{cost} of a answer is its 
depth in the search tree (i.e. the number of calls to relational definitions 
required to find them) \citep{seres1999algebra}. In the following example, the 
cost of an answer is equal to the length of the inner lists plus the length of 
the outer list. Noted that BFS produces answers in increasing order of cost.
\NOTE[MVC]{Here inner and outer are very confusing. Can you be more specified?}

@@

Balanced interleaving DFS (DFS$_\textrm{bi}$) has almost-fair \disj{} and unfair \conj{}. 
The implementation of DFS$_\textrm{bi}$ differs from DFS$_\textrm{i}$ in the \disj{} macro. We list the 
new \disj{} with its helpers in Fig.~\ref{balanced-disj}. The first helper 
function, \texttt{split}, takes a list of goals \texttt{ls} and a procedure 
\texttt{k}, partitions \texttt{ls} into two sub-lists of roughly equal length, 
and returns the application of \texttt{k} to the two sub-lists. \texttt{disj*} 
takes a non-empty list of goals \texttt{gs} and returns a goal. With the help 
of \texttt{split}, it essentially constructs a \emph{balanced} binary tree 
where leaves are elements of \texttt{gs} and nodes are \texttt{disj2}s, hence 
the name of this search strategy. In contrast, the \disj{} in DFS$_\textrm{i}$
constructs the binary tree with the same nodes but in the unbalanced form.

@@

\NOTE[MVC]{Though bag is well known, people rarely say ``bagging''. How about putting information in a bag, or something better?}
\NOTE[MVC]{What is the bagging information?}
As we mentioned in section 2.2, our BFS bag answers by their cost. The bagging 
information is recorded subtly -- the \texttt{car}s of a search space have cost 
0 (i.e. they are in the same bag), and the costs of answers in thunk are 
computed recursively then increased by one. It is even more subtle that
\texttt{append-inf/fair} and the \texttt{append-map-inf/fair} respects the cost 
information. We make these facts more obvious by changing the type of search 
space, modifying related function definitions, and introducing a few more 
functions.

@@

\subsection{comparison with the BFS of  \citep{seres1999algebra}}

In this section, we compare the pure subset of our optimized BFS with the BFS 
found in \citep{seres1999algebra}. We focus on the pure subset because 
Silvija's system is pure. Their system represents search spaces with streams of 
lists of answers, where each list is a bag.

To compare efficiency, we translate her Haskell code into Racket (See 
supplements for the translated code). The translation is direct 
due to the similarity in both logic programming systems and search space 
representations. The translated code is longer and slower. Details about 
difference in efficiency are in section 6.

\NOTE[MVC]{This is way too much into Coq. What we want to say here is that
  for any number n and goal g, (run n g) our system generates the same result
as Seres' does.}
We prove the two search strategies are equivalent in Coq. Since search space 
can be infinite, we should use a co-inductive data type. However, Coq is too 
strict in the guardedness condition to accept a direct translation of the 
implementations. Therefore, we prove core theorems with finite search space 
instead. In order to generalize the conclusion to the cases with infinite 
search space, we prove a few more theorems saying that whenever we query 
answers lower than some finite cost, we can restrict goals to truncate search 
spaces at some finite depth without changing the query result. (See supplements 
for the formal proof)

@@

In this section, we compare the efficiency of search strategies. A concise 
description is in Table~\ref{compare-efficiency}. A hyphen means running out of 
memory. The first two benchmarks are taken from 
\citet{friedman_reasoned_2018}. \texttt{reverso} is from 
\citet{rozplokhas2018improving}. Next two benchmarks 
about quine are modified from a similar test case in \citep{byrd2017unified}. 
The modifications are made 
to circumvent the need for symbolic constraints (e.g. $\neq$, 
\texttt{absento}). Our version generates de 
Bruijnized expressions and prevent closures getting into list. The two 
benchmarks differ in the \conde clause order of their relational interpreters. 
The last two 
benchmarks are about synthesizing expressions that evaluate to \texttt{'(I love 
you)}. This benchmark is also inspired by \citep{byrd2017unified}. Again, the 
sibling benchmarks differ in the \conde clause order of their relational 
interpreters. The first one 
has elimination rules (i.e. application, \texttt{car}, and \texttt{cdr}) at the 
end, while the other has them at the beginning. We conjecture that DFS$_\textrm{i}$ would 
perform badly in the second case because elimination rules complicate the 
problem when running backward. The evaluation supports our conjecture.
